---
title: "Assignment4"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(gapminder)
suppressPackageStartupMessages(library(tidyverse))
```


## General data reshaping and relationship to aggregation


## Join, merge, look up (Activity #1)

In the last assignment I had wanted to superimpose a line plot corresponding to the weighted mean of the life expectancy on top of a graph of life expectancy. I didn't know how to do it at the time, but I bet I can use a join function to do it now!

I will use `full_join()` to supplement my data in __Step 1__ and join with gapminder data __Step 2__. In __Step 3__ I plot the results, while in __Step 4__ I explore other `_join()` functions (`left_join()`, `anti_join()`).

### Step 1: create complementary dataframe

The first step is to create a dataframe that lists the weighted mean life expectancy for each continent and year:

```{r}
weightLifeExp <- gapminder %>%
  mutate(yearsLived = lifeExp*pop) %>% 
  group_by(year,continent) %>% 
  summarise(meanLifeExp = sum(as.numeric(yearsLived))
            /sum(as.numeric(pop)))

knitr::kable(head(weightLifeExp))
```

For fun, I will also include (via a `full_join` function!) Antarctica with life expectancy *NA* in the dataset (to make the different types of joins non-trivial).

```{r}
# create Antarctica
Antarctica <- gapminder %>% 
  filter(country == "Canada") %>% 
  select(year) %>% # Just get a sequence of the years we have data for
  mutate(continent = factor('Antarctica'), penguins = floor(10^(8*2000/jitter(year))/jitter(year)))

# integrate Antarctica into the data
weightLifeExp <- weightLifeExp %>% 
  full_join(Antarctica) %>% 
  arrange(year, continent)

head(weightLifeExp)
```



### Step 2: join data frames

Now we can join the two dataframes using `_join` functions (I also remove the GDP column for compactness):
```{r}
meanlifegap <- gapminder %>% 
  full_join(weightLifeExp) %>% 
  select(-gdpPercap)

head(meanlifegap)
```


### Step 3: make plot

Not one of the steps suggested by the assignment, but I personally wanted to redo the plot from my last assignment and see how the weighted mean compared to the `loess` trend line:

```{r}
ggplot(meanlifegap, aes(year, lifeExp)) + facet_wrap(~ continent) +
  geom_smooth(aes(color = 'blue'), method = loess, show.legend = T) +
  geom_point(alpha = 0.1) + 
  geom_line(aes(x = year, y = meanLifeExp, color = 'red'), show.legend = T) +
  labs(title = 'Life expectancy on each continent', y = 'Life Expectancy') +
  scale_color_discrete(name = "Trend Lines", labels = c("Loess trend line", "Weighted mean")) +
  theme(plot.title = element_text(hjust = 0.5))
```

We can see that they generally agree, but diverge pretty badly in the Americas and Asia at the beginning of the dataset. We can also see that the use of `full_join()` created an empty Antarctica graph (because all of those values are *NA*), and gave us the error message. 

### Step 4: explore different types of joins

We can quickly double check that Antarctica is the only continent unrepresented in the gapminder database using `inner_join()`:

```{r}
anti_join(weightLifeExp, gapminder)

anti_join(gapminder, weightLifeExp)
```

This shows us that the only rows in `weightLifeExp` with continent and year not in `gapminder` are Antarctica, while there are no rows in `gapminder` with either continent or year not in `weightLifeExp`.

If we wanted to get rid of the empty antarctica plot we could use `left_join()`:

```{r}
meanlifegap <- left_join(gapminder, weightLifeExp)

ggplot(meanlifegap, aes(year, lifeExp)) + facet_wrap(~ continent) +
  geom_smooth(aes(color = 'blue'), method = loess, show.legend = T) +
  geom_point(alpha = 0.1) + 
  geom_line(aes(x = year, y = meanLifeExp, color = 'red'), show.legend = T) +
  labs(title = 'Life expectancy on each continent', y = 'Life Expectancy') +
  scale_color_discrete(name = "Trend Lines", labels = c("Loess trend line", "Weighted mean")) +
  theme(plot.title = element_text(hjust = 0.5))
```

To explore the rest of the functions I'm going to trucate data quite a bit by considering only Canadian data after the centennial:

```{r}
Canada100 <- gapminder %>%
  filter(country == 'Canada', year >= 1967)
```

We can see what data in `weightLifeExp` corresponds to this new dataset by using the `semi_join()` function:

```{r}
semi_join(weightLifeExp, Canada100)
```

Likewise, if we want to see all the data we have corresponding to this dataset, we can use the `inner_join()` function:

```{r}
inner_join(weightLifeExp, Canada100)
```

### Step 5 (Activity #3): `merge()` and `match()`

